Modern Statistical Methods --- Example Sheet 1

### 2020-10-20 inverse_problems/C1.tex
worked on msm exercise sheet and almost finished inverse problems notes up to lecture 6

\section{Generalised Solutions}
\begin{recap}
    \begin{enumerate}
        \item     An linear operator $A \colon \XX \to \YY$ is called \emph{bounded} if
        \[
        \norm{A}_{\LL(\XX, \YY)} \ceq \sup_{u \neq 0} \frac{\norm{Au}_\YY}{\norm{u}_\XX} = \sup_{\norm{u}_\XX \leq 1} < \infty. 
        \]
            It is known that a linear operator between normed spaces is continuous if and only if it is bounded. The set of bounded linear operators from $\XX$ to $\YY$ is denotes $\LL(\XX, \YY)$. 
            \item 
        We let $\Dom(A)$, $\Nul(A)$ and  $\Ran(A)$ denote the domain, null space, and range of $A$ respectively.

        \item     We will assume $\XX$ and $\YY$ are Hilbert spaces, so there is an inner product $\ang{\cdot, \cdot}$ and any bounded operator $A \in \LL(\XX, \YY)$ has a unique adjoint $A^* \in \LL(\YY, \XX)$ which satisfies
        \[
        \ang{Au, v}_\YY = \ang{u, A^* v}_\XX \quad\text{for all $u \in \XX$, $v \in \YY$.}
        \]

@@

    \item For any $A \in \LL(\XX, \YY)$ we have
    \[
    \Ran(A)\P = \Nul(A^*) \quad\text{and}\quad \Nul(A)\P = \clos{\Ran(A^*)}. 
    \]

@@

\begin{lemma}
    For any $A \in \LL(\XX, \YY)$ we have $\clos{\Ran(A^*A)} = \clos{\Ran(A^*)}$. 
\end{lemma}

@@

\subsection{Generalised inverses}
We consider the equation \begin{equation} \label{eq:basic_ip}
    Au = f,
\end{equation} $A \in \LL(\XX, \YY)$, $f$ is known, and we wish to find $u$. 
\begin{definition}
    An element $u \in \XX$ is called a \emph{least-squares solution} of \cref{eq:basic_ip} if $u$ is a minimiser of the function $v \mapsto \norm{Av - f}_\YY$. It is called a \emph{minimal-norm solution} of \cref{eq:basic_ip} if it has minimal norm among all least-squares solutions. 
\end{definition}

@@

\begin{theorem}
    Let $f \in \YY$ and $A \in \LL(\XX, \YY)$. Then the following are equivalent:
    \begin{enumerate}
        \item $u \in \XX$ satisfies $Au = P_{\clos{\Ran(A)} }f$;
        \item $u$ is a least-squares solution of $\cref{eq:basic_ip}$:
        \item $u$ solves the \uline{normal equation}
        \begin{equation} \label{eq:normal}
                   A^* f = A^* A u. 
        \end{equation}
    \end{enumerate}
\end{theorem}

@@

\begin{definition}
    Let $A \in \LL(\XX, \YY)$, and define
    \[
    \tilde A \ceq A \restr_{\Nul(A)\P} \colon \Nul(A)\P \to \Ran(A). 
    \]
    Clearly $\tilde A$ is bijective and we define the \emph{Moore-Penrose inverse} 
    \[
    A\D \colon \Ran(A) \oplus \Ran(A)\P \to \Nul(A)\P \colon f \mapsto \tilde A^{-1} P_{\clos{\Ran(A)}} f. 
    \]
\end{definition}

@@

\subsection{Compact operators}
\begin{definition}
    Let $A \in \LL(\XX, \YY)$. Then $A$ is called \emph{compact} if for any bounded $B \subseteq \XX$, the image $A(B)$ is precompact in $\YY$. The set of compact operators in $\LL(\XX, \YY)$ is denoted $\KK(\XX, \YY)$. 
\end{definition}

\begin{lemma}
    Let $A \in \LL(\XX, \YY)$. Then $A$ is compact if and only if, for every bounded sequence $(x_n) \subseteq X$, the sequence $(Ax_n) \subseteq Y$ has a convergent subsequence. 
\end{lemma}

>>>>

\section{Generalised Solutions}
\begin{recap}
    \begin{enumerate}
        \item     An linear operator $A \colon \XX \to \YY$ is called \emph{bounded} if
        \[
        \norm{A}_{\BB(\XX, \YY)} \ceq \sup_{u \neq 0} \frac{\norm{Au}_\YY}{\norm{u}_\XX} = \sup_{\norm{u}_\XX \leq 1} < \infty. 
        \]
            It is known that a linear operator between normed spaces is continuous if and only if it is bounded. The set of bounded linear operators from $\XX$ to $\YY$ is denotes $\BB(\XX, \YY)$. 
            \item 
        We let $\Dom(A)$, $\Nul(A)$ and  $\Ran(A)$ denote the domain, null space, and range of $A$ respectively.

        \item     We will assume $\XX$ and $\YY$ are Hilbert spaces, so there is an inner product $\ang{\cdot, \cdot}$ and any bounded operator $A \in \BB(\XX, \YY)$ has a unique adjoint $A^* \in \BB(\YY, \XX)$ which satisfies
        \[
        \ang{Au, v}_\YY = \ang{u, A^* v}_\XX \quad\text{for all $u \in \XX$, $v \in \YY$.}
        \]

@@

    \item For any $A \in \BB(\XX, \YY)$ we have
    \[
    \Ran(A)\P = \Nul(A^*) \quad\text{and}\quad \Nul(A)\P = \clos{\Ran(A^*)}. 
    \]

@@

\begin{lemma}
    For any $A \in \BB(\XX, \YY)$ we have $\clos{\Ran(A^*A)} = \clos{\Ran(A^*)}$. 
\end{lemma}

@@

\subsection{Generalised inverses}
We consider the equation \begin{equation} \label{eq:basic_ip}
    Au = f,
\end{equation} where $A \in \BB(\XX, \YY)$ and $f$ are known, and we wish to find $u$. 
\begin{definition}
    An element $u \in \XX$ is called a \emph{least-squares solution} of \cref{eq:basic_ip} if $u$ is a minimiser of the function $v \mapsto \norm{Av - f}_\YY$. It is called a \emph{minimal-norm solution} of \cref{eq:basic_ip} if it has minimal norm among all least-squares solutions. 
\end{definition}

@@

\begin{theorem}
    Let $f \in \YY$ and $A \in \BB(\XX, \YY)$. Then the following are equivalent:
    \begin{enumerate}
        \item $u \in \XX$ satisfies $Au = P_{\clos{\Ran(A)} }f$;
        \item $u$ is a least-squares solution of $\cref{eq:basic_ip}$:
        \item $u$ solves the \uline{normal equation}
        \begin{equation} \label{eq:normal}
                   A^* f = A^* A u. 
        \end{equation}
    \end{enumerate}
\end{theorem}

@@

\begin{definition}
    Let $A \in \BB(\XX, \YY)$, and define
    \[
    \tilde A \ceq A \restr_{\Nul(A)\P} \colon \Nul(A)\P \to \Ran(A). 
    \]
    Clearly $\tilde A$ is bijective and we define the \emph{Moore-Penrose inverse} 
    \[
    A\D \colon \Ran(A) \oplus \Ran(A)\P \to \Nul(A)\P \colon f \mapsto \tilde A^{-1} P_{\clos{\Ran(A)}} f. 
    \]
\end{definition}

@@

\subsection{Compact operators}
\begin{definition}
    Let $A \in \BB(\XX, \YY)$. Then $A$ is called \emph{compact} if for any bounded $B \subseteq \XX$, the image $A(B)$ is precompact in $\YY$. The set of compact operators in $\BB(\XX, \YY)$ is denoted $\KK(\XX, \YY)$. 
\end{definition}

\begin{lemma}
    Let $A \in \BB(\XX, \YY)$. Then $A$ is compact if and only if, for every bounded sequence $(x_n) \subseteq X$, the sequence $(Ax_n) \subseteq Y$ has a convergent subsequence. 
\end{lemma}

### 2020-10-20 inverse_problems/C2.tex
updatd msm & ip notes
worked on msm exercise sheet and almost finished inverse problems notes up to lecture 6

\section{Classical regularisation theory}
\begin{definition}
    Let $A \in \LLL(\XXX, \YYY)$ and 
\end{definition}

>>>>

\section{Classical regularisation theory}
Let $A \in \BB(\XX, \YY)$ such that $\Ran(A)$ is not closed (this happens for example when $A$ is compact and does not have finite rank), and consider the inverse problem $Au = f$. Suppose we measure not $f$, but noisy data $f_\delta$ such that $\norm{f_\delta - f} \leq \delta$. Then since $A\D$ is discontinuous, we cannot expect that $A\D f_\delta \to A\D f$ as $\delta \to 0$. Therefore, we must replace $A\D$ by operators that approximate it.

\begin{definition}
	Let $A\in \BB(\XX, \YY)$. A family $(R_\alpha)_{\alpha > 0}$ of continuous operators is called a \emph{regularisation} of $A\D$ if
	\[
	\lim_{\alpha \to 0} R_\alpha f = A\D f \quad\text{for all $f \in \Dom(A\D)$}. 
	\]
	If all $R_\alpha$ are linear, then we speak of a \emph{linear regularisation} of $A\D$. 
\end{definition}

\begin{theorem}[Banach-Steinhaus]
	Let $\XX, \YY$ be Hilbert spaces and $\qty{A_\alpha} \subseteq \BB(\XX, \YY)$ a family of pointwise bounded operators. Then $\qty{A_\alpha}$ is bounded in norm. 
\end{theorem}

\begin{corollary} \label{cor:bounded_converge}
	Let $\XX, \YY$ be Hilbert spaces and $(A_j) \subseteq \BB(\XX, \YY)$. Then $(A_j)$ converges pointwise to some $A \in \BB(\XX, \YY)$ if and only if $\qty{A_j}$ is norm-bounded and converges pointwise on some dense subset $\XX' \subseteq \XX$. 
\end{corollary}

\begin{theorem}
	Let $\XX, \YY$ be Hilbert spaces, $A \in \BB(\XX, \YY)$ and $(R_\alpha)_{\alpha > 0}$ a linear regularisation. If $A\D$ is not continuous, $(R_\alpha)$ is not norm-bounded. In particular, there exists $f \in \YY$ with $\norm{R_\alpha f} \to \infty$. 
\end{theorem}

\begin{proof}
	Suppose $(R_\alpha)$ is norm-bounded. Let $\alpha_j \to 0$, then we know that $R_{\alpha_j} \to A\D$ pointwise on $\Dom(A\D)$. Since $\Dom(A\D)$ is dense in $\YY$, \cref{cor:bounded_converge} then  tells us that $A\D$ is bounded and therefore continuous, a contradiction.

	By the Banach-Steinhaus theorem, if $(R_\alpha)$ is not norm-bounded, it is not pointwise bounded, so there must exist $f \in \YY$ such that $\qty{\norm{R_\alpha f}}$ is not bounded. 
\end{proof}

\begin{recap}
	Recall that any bounded sequence in a Hilbert space has a weakly convergent subsequence. 
\end{recap}
\begin{theorem}
	Let $A \in \BB(\XX, \YY)$ and $(R_\alpha)$ a linear regularisation of $A\D$. If $\qty{\norm{A R_\alpha}}_{\alpha > 0}$ is bounded, then $\norm{R_\alpha f} \to \infty$ as $\alpha \to 0$ for every $f \notin \Dom(A\D)$. 
\end{theorem}

\begin{proof}
	Define $u_\alpha \ceq R_\alpha f$ for $f \notin \Dom(A\D)$, and assume there exists a sequence $\alpha_k \to 0$ such that $\qty{\norm{u_{\alpha_k}}}$ is bounded. After taking a subsequence if necessary, we may assume that $u_{\alpha_k} \rightharpoonup u$ for some $u \in \XX$, and therefore we also have $Au_{\alpha_k} \rightharpoonup Au$.

	We also have $\lim_{\alpha \to 0} A R_\alpha f = AA\D f = P_{\clos{\Ran(A)}}f$ for $f \in \Dom(A\D)$, and since we assumed $\qty{A R_\alpha}$ was norm-bounded, by \cref{cor:bounded_converge} we have $\lim_{\alpha \to 0} AR_\alpha f = P_{\clos{\Ran(A)}} f$ for all $f \in \YY$.

	Since $Au_{\alpha_k}$ is convergent and has weak limit $Au$, it must also have limit $Au$, so we find $Au = P_{\clos{\Ran(A)}} f$ so $f \in \Dom(A\D)$, a contradiction. 
\end{proof}

We need some process to choose a parameter. To this end, note that we have
\[
\norm{R_\alpha f_\delta - A\D f} \leq \norm{R_\alpha (f_\delta - f)} + \norm{(R_\alpha - A\D) f} \\
\leq  \norm{R_\alpha} \norm{f_\delta - f} + \norm{f} \norm{R_\alpha - A\D}. 
\]
The first term is called the \emph{data error} and is unbounded for $\alpha \to 0$, and the second term is called the \emph{approximation error} which does vanish for $\alpha \to 0$. Therefore, we want to choose $\alpha$ small enough to have a low approximation error, while keeping the data error at bay. We define this formally:

\begin{definition}
	A function $\alpha \colon  \RR_{> 0} \times \YY \to \RR_{> 0} \colon (\delta, f_\delta) \mapsto \alpha(\delta, f_\delta)$ is called a \emph{parameter choice rule} (PCR). We distinguish three types:
	\begin{enumerate}
		\item An \emph{a priori} PCR depends only on $\delta$;
		\item An \emph{a posteriori} PCR depends on both $\delta$ and $f_\delta$;
		\item A \emph{heuristic} PCR depends only on $f_\delta$. 
	\end{enumerate}
\end{definition}

### 2020-10-20 modern_statistical_methods/msm_ES1.tex
worked on msm exercise sheet and almost finished inverse problems notes up to lecture 6

\begin{solution}
	Let $\XX = [0, 1]$ and $k(x, x') = \min(x, x')$ the Sobolev kernel. We must then show that for any $x_1, \dotsc, x_n \in \XX$, the matrix $K_{ij} = \min(x_i, x_j)$ is positive semi-definite. 
\end{solution}
\end{document}

>>>>

\begin{solution}
	Let $\XX = [0, 1]$ and $k(x, x') = \min(x, x')$ the Sobolev kernel. We must then show that for any $x_1, \dotsc, x_n \in \XX$, the matrix $K_{ij} = \min(x_i, x_j)$ is positive semi-definite (symmetry of $k$ is clear).

	To this end, assume $x_1 \leq \dotsb \leq x_n$ (this may require relabeling of the $x_i$ which does not affect positive semi-definiteness of $K$). Now let $\vec f_i$ denote the vector $f_{i1}, \dotsc, f_{i(i-1)} = 0$, $f_{ii}, \dotsc, f_{in} = 1$. 
	\[
	K = \mqty[ x_1 & x_1 & x_1 & \cdots & x_1 \\
	x_1 & x_2 & x_2 & \dotsb & x_2 \\
	x_1 & x_2 & x_3 & \dotsb & x_3 \\
	\vdots & \vdots & \vdots & \ddots & \vdots \\
	x_1 & x_2 & x_3 & \cdots & x_n] = x_1 \vec f_1 \vec f_1\T + (x_2 - x_1) \vec f_2 \vec f_2\T + \dotsb + (x_n - x_{n-1}) \vec f_n \vec f_n\T. 
	\]
	Since this is a nonnegative combination of positive semidefinite matrices, we conclude that $K$ is positive semidefinite. 
\end{solution}
\end{document}
